Boosting Higgs Pair Production in the $bb\bar{b}\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-07 paper/sec-mva.tex
working on MVA section

 \subsection{Input kinematical variables for the MVA}

@@

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in $R$ between the two Higgs candidates, $\Delta R_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item In the boosted category,
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,1}}$, Higgs candidate subjets.
  %
  In the resolved category instead, we use the transverse momenta $p_{T,i}$ of the four leading 
   $b$-tagged small-$R$ jets in the event. 
\end{itemize}
Therefore, we have a total of 13 variables which are common to the three categories.

@@

In order to gain physical insight of which are the variables that
contribute the most to the MVA discrimination it is useful
to show the distribution of NN weights for each of the
input variables.
%
This distributions is a measure of the correlation between
each input variable and the Neural Network output: the higher
the value, the stronger the impact of this variable in
separating signal from the background.
%
In Fig.~\ref{fig:nnweights} we show
the distribution of NN weights for the resolved and boosted categories.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.48\textwidth]{plots/nnweights_res.pdf}
\includegraphics[width=0.48\textwidth]{plots/nnweights_boost.pdf}
\caption{\small Distribution of weights per input variable in the final ANN analysis. The above figure shows the weight distribution in the case of the resolved analysis, while the lower plot shows the boosted analysis.}
\label{fig:nnweights}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

The results of Fig.~\ref{fig:nnweights} clearly illustrate which variables
are more discriminative between the signal and the background.
%
For instance, in the case of the resolved analysis, the variables that show
a higher discrimination power is the invariant mass of the Higgs candidate dijets.
%
Note that as explained in Sect.~\ref{sec:analysis}, we have applied a realistic
smearing of the four-momentum of all reconstructed jets, but still is clear that the
distribution of invariant mass of the Higgs candidates will be very different
in the signal and in the backgrounds.
%
We discussed below the impact of smearing has.
%
Other variables do not seem to have a dramatic effect, although of course one of the
advantages of the MVA technique is that it does not hurt to add redundant
information.
%
In particular the angular distances between the four leading $b$-jets does not
provide additional information.

The corresponding analysis for the boosted category is also
shown in Fig.~\ref{fig:nnweights}.
%
here we see that there is a more even distributions of weights, in that many
variables help in a similar amount in the discrimination.
%
The invariant mass of the Higgs candidate of course still carries a lot of weight,
but now we see that the substructure variables really carry  a lot of information,
in particular the splitting scales and the subjetiness $\tau_{12}$.
%
We have verified that signal significance in the boosted category is substantially
degraded if substructure variables are not used.

@@

\subsection{Impact of smearing in the MVA discrimination}

Accounting for a suitable smearing of the final-state momenta of the reconstructed
jets is essential to be able to draw any realistic conclusions in analysis like
this one that are based on a large number of jets.
%
It is therefore interesting how the MVA discrimination changes if one did not account
for smearing - of course in this case differences between signal and background distributions
would be enhanced.

To investigate this, we now consider a variant of the analysis used in
Table  where the $m_h$ window criterion has been relaxed to $80 < m_h < 170$.
%
Firstly the same analysis is repeated with the larger $m_h$ window, with no smearing at all.
%
This is then compared to the same analysis but with a Gaussian smearing with $\sigma=10$ GeV applied to the dijet invariant mass used in the MVA.
%
In the left panel of Fig.~\ref{fig:invMsmear} the smearing is demonstrated.
%
Both with and without smearing, the  invariant mass distribution of the
Higgs candidates is still the most discriminative variable.
%
This can be seen by comparing with   with the corresponding distribution with smearing in
  Fig.~\ref{fig:nnweights}.

%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.49\textwidth]{plots/Msmear/histo_mass_smear_precut.pdf}
\includegraphics[width=0.49\textwidth]{plots/Msmear/nnweights.pdf}
\caption{\small Left plot: the invariant mass distribution of the
  Higgs candidates without (red) and with (blue) smearing.
  %
  Right plot: the strength of the NN weights in the resolved case when no smearing
  is applied.
  %
  It should be compared with the corresponding distribution with smearing in
  Fig.~\ref{fig:nnweights}.
  %
  As expected, the discrimination power of the invariant mass distribution is
  enhanced if there is no smearing.}
\label{fig:invMsmear}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%

While the MVA still derives most of it's discriminating power from the dijet masses, the smearing reduces it's effectiveness considerably.
%
In Fig.~\ref{fig:smearedROC} the ROC curves of the MVA applied to the smeared and unsmeared distributions are shown, where it is clear that the overall performance of the MVA is significantly hampered by the smeared invariant masses.
%
For example, if we require the same signal efficiency as before, 0.8, the background rejection
is degraded from 80\% down to 60\%.
%
Therefore, we can conclude that it is important to include smearing to draw realistic
conclusions, though the qualitative information of which are the most relevant
discriminative variables is not changed by this conclusion.

%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.50\textwidth]{plots/Msmear/roc.pdf}
\caption{\small ROC curves of the MVA applied with the unsmeared (blue) and smeared (green) dijet invariant mass distributions.}
\label{fig:smearedROC}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%

>>>>

 \subsection{Input kinematical variables for the MVA}
 \label{sec:input}

@@

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in $R$ between the two Higgs candidates, $\Delta R_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item In the boosted category,
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,1}}$, Higgs candidate subjets.
  %
  In the resolved category instead, we use the transverse momenta $p_{T,i}$ of the four leading 
   $b$-tagged small-$R$ jets in the event. 
\end{itemize}
Therefore, we have a total of 13 variables which are common to the three categories.
%

@@

A very useful property of MVA's such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the signal over background
separation.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^5 \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will those
be the ones that play a dominant role in enhancing the signal
discrimination using the MVA.

%
In Fig.~\ref{fig:nnweights} we show
the distribution of the total associated weight,
Eq.~(\ref{eq:totweight}) for each of the $N_{\rm var}$ input
variables of the resolved (left plot) and boosted (right plot) categories.
%
The notation for the various kinematic variables is the same
as in Sect.~\ref{sec:input}.
%
The scale of the $y$ axis is arbitrary, and the important information
is contained in the relative strengths of the total associated weight
for each of the input variables.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.48\textwidth]{plots/res_wgthist.pdf}
\includegraphics[width=0.48\textwidth]{plots/boosted_wgthist.pdf}
\caption{\small
Distribution of the total associated weight,
Eq.~(\ref{eq:totweight}) for each of the $N_{\rm var}$ input
variables of the resolved (left plot) and boosted (right plot) categories.
%
See Sect.~\ref{sec:input} for a description of each of the input
variables to the MVA in the two cases.
}
\label{fig:nnweights}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

The results of Fig.~\ref{fig:nnweights} clearly illustrate which variables
are more discriminative between the signal and the background.
%
Let's begin discussing the resolved analysis.
%
In this category, the variables that lead to
a higher discrimination power
are the $p_T$ of the two reconstructed Higgs, and then on a similar
footing other variables like the $p_T$ of the individual jets
and the mass of the di-Higgs system.
%
The only variable which seems to contain to information at all
is the $\Delta \phi_{hh}$ separation between the two
Higgs candidates.

For the boosted category, we see that substructure variables
are the most helpful ones, in particular the $k_t$ splitting scales
$\sqrt{d_{12}}$ and the subjetiness ratio $\tau_{12}$.
%
The ratio of energy correlations $D_2$ and $C_2$ also help,
in particular the former.
%
From the more traditional variables, the angular
separation between the two large-$R$ jets $\Delta R_{hh}$,
the invariant mass of the di-Higgs system $m_{hh}$ and the various
$p_T$ also have discrimination power.
%
Other variables provide limited information, in particular the invariant
mass of the Higgs candidates, $m_{h_1}$ and $m_{h_2}$, due to the realistic
simulation of detector resolution introduced with the smearing.
%
We have verified that signal significance in the boosted category is substantially
degraded if substructure variables are not used, emphasizing that
in the boosted category jet substructure provides the best handle
to separate signal and background.

At this point we have all the ingredients that we need to determine the optimal
value of the MVA output cut and quantify the final values of the signal
significance $S/\sqrt{B}$ and $S/B$ in our three categories.
%
These optimal values are determining by the maximisation of $S/\sqrt{B}$,
ensuring that the number of events which is left at the HL-LHC is large
enough, and also that the actual number of MC events used for the ANN training
is still large enough to avoid the biases of a small training sample.
%
The post-MVA results for $S/\sqrt{B}$ and $S/B$ as a function of the cut
in the ANN output for each of the three categories are shown in
Fig.~\ref{fig:sb_mva}.
%
As mentioned before, the optimal cuts are determined independently for each
category.
%
Note that the values of $S/\sqrt{B}$ and $S/B$ in the three cases should correspond
to the same values at the end of the cut-based analysis, as
can be verified by looking at the {\bf C2} row of Table~\ref{table:cutflow}.
%
Indeed, at the end of the cut based analysis $S/\sqrt{B}$ is approximately
1.0, 0.5 and 0.5 for the boosted, intermediate and resolved
categories, as can be seen from Fig.~\ref{fig:sb_mva}.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.48\textwidth]{plots/ssb.pdf}
\includegraphics[width=0.48\textwidth]{plots/sb.pdf}
\caption{\small
  The values of the signal significance $S/\sqrt{B}$ and of the
  signal over background ratio $S/B$ for the boosted, intermediate
  and resolved categories.
  %
  Note that the optimal cuts are determined independently for each
  category.
  %
  The no-cut values correspond to the  cut-based analysis
  results from Table~\ref{table:cutflow}.
}
\label{fig:sb_mva}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

The results of Fig.~\ref{fig:sb_mva} are interesting in that they show
that for the resolved and boosted categories the improvement in signal
significance arising from the MVA is rather moderate, less that a factor two
in the two cases.
%
For no value of the cut the $S/\sqrt{B}$ value in these two categories
goes above one.
%
The real usefulness of the MVA can be seen for the resolved category,
where the signal significance rises from around 1.0 in the cut-based
analysis to above 3.0 for values of the ANN output cut above 0.85.
%
This value of the cut seems reasonable, since we know from Fig.~\ref{fig:nev2}
that the number of signal events at the HL-LHC is still quite substantial,
of the order of 100.
%
Therefore, thanks to the MVA, we predict that one should be able
to find evidence for the production of Higgs boson pairs
at the HL-LHC using the $4b$ channel alone.

Another very important benefit of the MVA applied to the boosted
category is that fact that the background is substantially
reduced so that $S/B$ can be as large as 10\% for the optimal
values of the ANN output cut.
%
Note that this is not true for the other two categories,
where $S/B$ is roughly constant.
%
A reasonably large value of $S/B$ is extremely beneficial
from the experimental point of view, since it indicates
that the measurement is feasible even if the
systematic uncertainties are not very small.

Our final results are summarized in Table, where we indicate
the value of the optimal cut in each category,
the number of signal and background events $N_{\rm ev}$,
and also $S/\sqrt{B}$ and $S/B$.
%
For completeness, we also include the results after the cut-based
analysis.
%
Given that the signal significance is much larger in the boosted as
compared to the other two categories, no additional information
could be obtained by combining the significance of the three
categories, and it seems more efficient to perform
the analysis exclusively in the boosted category.

@@
